### 2024-04-04 appendices/app_case_study.tex
Intro post-review DC

>>>>

\section{Monthly versus hourly pathway optimisation}
\label{app:mo_vs_td}

This section compares the perfect foresight optimisation based on a monthly time resolution with the reference case (hourly time resolution). The main advantage of the monthly approach is its computational time, \ie a reduction of 99.8\% compared to the hourly resolution (see Table \ref{tab:TD_MO}). This tractability represents a significant advantage when several thousands of runs are necessary, \eg for uncertainty quantification as it has been performed in \cite{rixhon2021role}.

\begin{table}[htbp]
\caption{Comparison between the two different time-resolutions: Hourly (PF) and Monthly (MO). Differences with the reference case (PF) below 1\% are not shown ($\simeq$) and ones above 10\% are in bold.}
\label{tab:TD_MO}
\begin{minipage}{\textwidth}
\centering
\resizebox{\textwidth}{!}{
\begin{tabular}{l l r r c}
\toprule
& & {\thead{\textbf{PF}}} & {\thead{\textbf{MO}}} & Units \\
\midrule
Computational time\footnote{\label{foot:comp_time_mo}These computational times were reached on a 2.4GHz 4-core machine.} & & 830 & \textbf{2} & s \\
\midrule
\multirow{4}{*}{Costs in 2050}
& Total transition\footnote{\label{foot:transition_cost_mo}As detailed in \autoref{eq:obj_func_v2}, the transition cost is the sum of the cumulative opex and capex, salvage value being deduced.} & 1004 & -2\% & b€\textsubscript{2015}\\
& Cumulative opex & 528 & -3\%  & b€\textsubscript{2015}\\
& Cumulative capex & 636 & -2\% & b€\textsubscript{2015}\\
& Salvage value & 160 & -5\% & b€\textsubscript{2015}\\
\midrule
\multirow{5}{*}{Primary energy mix in 2050} & Total & 368.0 & +2\% & TWh/y\\
 & e-hydrogen & 15.6 & +2\% & TWh/y\\
 & e-methane & 41.0 & \textbf{+51\%} & TWh/y\\
 & e-methanol & 54.8 & $\simeq$ & TWh/y\\
 & e-ammonia & 40.7 & \textbf{+55\%} & TWh/y\\

 \midrule
 \multirow{3}{*}{Electrification in 2050\footnote{\label{foot:share_elec_mo}The electrification of the other sectors (\ie centralised heat (100\%-heat pump), private (100\%-\gls{BEV}), public mobility (80\%-train and tramway), freight mobility (25\%-train) and non-energy demand (0\%)) are identical between the three approaches and are, therefore, not presented in the table.}} & System\footnote{\label{foot:system_elec_mo}The electrification of the system is computed as the difference between the total production of electricity and the end-use demand of electricity.} & 63.3 & \textbf{-27\%} & TWh$_{\text{e}}$\\ 
 & Industrial heat\footnote{\label{foot:heat_elec_mo}The electrification of the industrial and decentralised heating sectors are expressed in terms of thermal energy (TWh$_{\text{th}}$) provided by electrified processes, respectively industrial resistors and decentralised electric heat pumps.} & 12.3 & \textbf{-98\%} & TWh$_{\text{th}}$\\
 & Decentralised heat\footref{foot:heat_elec_mo} & 73.9 & \textbf{-13\%} & TWh$_{\text{th}}$\\
  \midrule
\multirow{3}{*}{Year of full VRES-deployment} & PV & 2045 & \textbf{2040} & -\\
 & Wind-offshore & 2030 & \textbf{2025} & -\\
 & Wind-onshore & 2025 & 2025 & -\\
\bottomrule
\end{tabular}
}
\end{minipage}
\end{table}

However, averaging time series of end-use demands and renewable productions brings some discrepancies. First and foremost, \gls{VRES} loose their intrinsic power intermittency \cite{haydt2011relevance}. This has a series of consequences. Overall, this overestimates the uptake and emergence of solar and wind power generation and, consequently, overestimate their share in the primary electricity mix, especially at intermediate stages of the transition \cite{haydt2011relevance,Poncelet2016} (see Figure \ref{fig:MO_Share_VRES_elec_mix}). In line with these studies, the more there is \gls{VRES} in the primary electricity mix, the higher is the error\footnote{Like \citet{Poncelet2016}, this error is defined as:$\sum_{i} \frac{|\text{supply share}_i^{MO}-\text{supply share}_i^{PF}|}{2}$, where index i runs over all technologies and import of electricity and the supply shares are expressed as a percentage.} with lower time resolution. In 2025 and 2050, where the share of \gls{VRES} in the electricity primary mix is 46\% and 66\%, respectively, this error goes from 10\% to 18\%.

 \begin{figure}[!htbp]
\centering
\includegraphics[width=\textwidth]{MO_Share_VRES_elec_mix.pdf}
\caption{Share of \gls{VRES} (\ie solar and wind) in the electricity primary mix for the monthly and hourly time resolutions.}
\label{fig:MO_Share_VRES_elec_mix}
\end{figure}

When considering the system overall, this averaged time-variability requires a more limited sector-coupling in the monthly model to absorb the intermittency of renewables. Industrial and decentralised heat demands are less electrified (see \autoref{tab:TD_MO}). This leads to 30\% electrification of the overall system versus 36\% for the hourly approach. This smaller production of electricity is done to the detriment of direct import of electricity, completely removed from the primary mix, as assumed to keep on having a higher global warming potential than conventional assets (\eg \gls{CCGT} running on gas and, later on, renewable gas and ammonia). Then, the generation mix in heating sectors or non-energy demands is less diversified, leading the sharper and later switch of resources and technologies. For instance, lignocellulosic biomass is ``cannibalized" by industrial boilers to the detriment of biomass-to-methanol or \gls{HVC}. Besides this, in general, we observe an overestimation of conventional technologies (\eg running on gas or oil) generation, either by installing more capacities or having a higher load factor for similar capacities. In line with \citet{Poncelet2016}, this leads to an overall underestimation of the operational (and investment) costs (see \autoref{tab:TD_MO}).

Finally, in the mobility sectors, there is no difference between low and high time resolutions. This is due to the fact that these demands are assumed to stay the same over the different hours of the year (\ie freight mobility) or have a favorite technology to supply them (\eg \gls{BEV} for private mobility).

### 2024-04-04 appendices/app_results_UQ.tex
Intro post-review DC

\begin{table}[htbp!]
\caption{Total Sobol' indices of the uncertain parameters over the total transition cost in the hourly and monthly pathway models. The similar rankings (and indices) show the validity of using the faster (even though less accurate) monthly model to assess uncertainties.}
\label{tab:UQ_full}
\begin{minipage}{\textwidth}
\centering
\resizebox{0.8\textwidth}{!}{
\begin{tabular}{l c c|c c}
\toprule
\multirow{2}{*}{\textbf{Parameter}}  & \multicolumn{4}{c}{\textbf{Ranking (Sobol' index)}}\\
& \multicolumn{2}{c|}{Hourly model} & \multicolumn{2}{c}{Monthly model}\\ 	
\midrule
\textbf{Purchase electrofuels} & 1 & (46.8\%) & 1 & (47.4\%) \\
Industry EUD & 2 & (23.2\%) & 2 & (23.5\%) \\
Interest rate & 3 & (12.0\%) & 3 & (11.0\%) \\
Purchase fossil fuels & 4 & (5.7\%) & 4 & (6.9\%) \\
\midrule
Variable OPEX of technologies & 5 & (3.1\%) & 5 & (2.9\%)\\
Purchase biofuels & 6 & (2.6\%) & 6 & (2.6\%) \\
CAPEX electric motor & 7 & (2.1\%) & 8 & (1.9\%) \\
Purchase electricity & 8 & (1.5\%) & 34 & ($<$0.1\%) \\
Hourly load factor wind turbines & 9 & (1.1\%) & 9 & (1.3\%)\\
Hourly load factor PV & 10 & (1.1\%) & 7 & (1.9\%) \\
\textbf{Potential capacity \gls{SMR}} & 11 & (0.9\%) & 11 & (0.9\%)\\
CAPEX car & 12 & (0.8\%) & 13 & (0.7\%) \\
Available local biomass & 13 & (0.7\%) & 12 & (0.8\%) \\
Passenger mobility EUD & 14 & (0.7\%) & 14 & (0.7\%) \\
Modal share change LT-heat & 15 & (0.5\%) & 15 & (0.5\%)\\
Max capacity PV & 16 & (0.5\%) & 10 & (1.1\%)\\
Households EUD & 17 & (0.5\%) & 16 & (0.5\%) \\
Services EUD & 18 & (0.5\%) &  17 & (0.4\%) \\
Max share of public transport & 19 & (0.3\%) & 19 & (0.3\%) \\
Max capacity onshore wind & 20 & (0.3\%) & 18 & (0.3\%) \\
CAPEX PV & 21 & (0.2\%) & 20 & (0.2\%) \\
Efficiency electric motor & 22 & (0.2\%) & 22 & (0.1\%)\\
Max capacity offshore wind & 23 & (0.2\%) & 21 & (0.2\%) \\
Available electricity import & 24 & (0.1\%) & 33 & ($<$0.1\%) \\
CAPEX ICE & 25 & (0.1\%) & 24 & (0.1\%)\\
CAPEX fuel cell engine & 26 & (0.1\%) & 23 & (0.1\%) \\
Efficiency fuel cell engine & 27 & (0.1\%) & 25 & ($<$0.1\%) \\
Modal share change freight mobility & 28 & (0.1\%) & 26 & ($<$0.1\%) \\
Modal share change passenger mobility & 29 & ($<$0.1\%) & 27 & ($<$0.1\%) \\
CAPEX grid reinforcement & 30 & ($<$0.1\%) & 30 & ($<$0.1\%) \\
CAPEX efficiency measures & 31 & ($<$0.1\%) & 28 & ($<$0.1\%) \\
CAPEX bus & 32 &($<$0.1\%)  & 29 & ($<$0.1\%) \\
\textbf{CAPEX \gls{SMR}} & 33 & ($<$0.1\%)  & 32 & ($<$0.1\%) \\
CAPEX power grid & 34 & ($<$0.1\%) & 31 & ($<$0.1\%) \\
\bottomrule

@@

Figures \ref{fig:results_uq_prod_cons_Gas}, \ref{fig:results_uq_prod_cons_H2}, \ref{fig:results_uq_prod_cons_Ammonia} and \ref{fig:results_uq_prod_cons_Methanol} give the distribution of the different routes of supply and consumption methane, hydrogen, ammonia and methanol, resulting from the 1260 samples of the \gls{GSA}.

>>>>

\begin{table}[htbp!]
\caption{Total Sobol' indices of the uncertain parameters over the total transition cost in the hourly and monthly (see Appendix \ref{app:mo_vs_td}) pathway models. The similar rankings (and indices) show the validity of using the faster (even though less accurate) monthly model to assess uncertainties.}
\label{tab:UQ_full}
\begin{minipage}{\textwidth}
\centering
\resizebox{0.8\textwidth}{!}{
\begin{tabular}{l c c|c c}
\toprule
\multirow{2}{*}{\textbf{Parameter}}  & \multicolumn{4}{c}{\textbf{Ranking (Sobol' index)}}\\
& \multicolumn{2}{c|}{Hourly model} & \multicolumn{2}{c}{Monthly model}\\ 	
\midrule
\textbf{Purchase electrofuels} & 1 & (46.8\%) & 1 & (47.4\%) \\
Industry EUD & 2 & (23.2\%) & 2 & (23.5\%) \\
Interest rate & 3 & (12.0\%) & 3 & (11.0\%) \\
Purchase fossil fuels & 4 & (5.7\%) & 4 & (6.9\%) \\
\midrule
Variable OPEX of technologies & 5 & (3.1\%) & 5 & (2.9\%)\\
Purchase biofuels & 6 & (2.6\%) & 6 & (2.6\%) \\
CAPEX electric motor & 7 & (2.1\%) & 8 & (1.9\%) \\
Purchase electricity & 8 & (1.5\%) & 34 & ($<$0.1\%) \\
Hourly load factor wind turbines & 9 & (1.1\%) & 9 & (1.3\%)\\
Hourly load factor PV & 10 & (1.1\%) & 7 & (1.9\%) \\
\textbf{Potential capacity \gls{SMR}} & 11 & (0.9\%) & 11 & (0.9\%)\\
CAPEX car & 12 & (0.8\%) & 13 & (0.7\%) \\
Available local biomass & 13 & (0.7\%) & 12 & (0.8\%) \\
Passenger mobility EUD & 14 & (0.7\%) & 14 & (0.7\%) \\
Modal share change LT-heat & 15 & (0.5\%) & 15 & (0.5\%)\\
Max capacity PV & 16 & (0.5\%) & 10 & (1.1\%)\\
Households EUD & 17 & (0.5\%) & 16 & (0.5\%) \\
Services EUD & 18 & (0.5\%) &  17 & (0.4\%) \\
Max share of public transport & 19 & (0.3\%) & 19 & (0.3\%) \\
Max capacity onshore wind & 20 & (0.3\%) & 18 & (0.3\%) \\
CAPEX PV & 21 & (0.2\%) & 20 & (0.2\%) \\
Efficiency electric motor & 22 & (0.2\%) & 22 & (0.1\%)\\
Max capacity offshore wind & 23 & (0.2\%) & 21 & (0.2\%) \\
Available electricity import & 24 & (0.1\%) & 33 & ($<$0.1\%) \\
CAPEX ICE & 25 & (0.1\%) & 24 & (0.1\%)\\
CAPEX fuel cell engine & 26 & (0.1\%) & 23 & (0.1\%) \\
Efficiency fuel cell engine & 27 & (0.1\%) & 25 & ($<$0.1\%) \\
Modal share change freight mobility & 28 & (0.1\%) & 26 & ($<$0.1\%) \\
Modal share change passenger mobility & 29 & ($<$0.1\%) & 27 & ($<$0.1\%) \\
CAPEX grid reinforcement & 30 & ($<$0.1\%) & 30 & ($<$0.1\%) \\
CAPEX efficiency measures & 31 & ($<$0.1\%) & 28 & ($<$0.1\%) \\
CAPEX bus & 32 &($<$0.1\%)  & 29 & ($<$0.1\%) \\
\textbf{CAPEX \gls{SMR}} & 33 & ($<$0.1\%)  & 32 & ($<$0.1\%) \\
CAPEX power grid & 34 & ($<$0.1\%) & 31 & ($<$0.1\%) \\
\bottomrule

@@

Figures \ref{fig:results_uq_prod_cons_Gas}, \ref{fig:results_uq_prod_cons_H2}, \ref{fig:results_uq_prod_cons_Ammonia} and \ref{fig:results_uq_prod_cons_Methanol} give the distribution of the different routes of supply and consumption of methane, hydrogen, ammonia and methanol, resulting from the 1260 samples of the \gls{GSA}.

### 2024-04-04 chap_RL/chap_RL.tex
Intro post-review DC

Before testing the optimal policy $\pi^*\left(\bm{a}_n | \bm{o}_n\right)$, the first step consists in assessing the training of the \gls{NN}. For this, numerous episodes are played through the myopic optimization of the transition pathway of Belgium. At the beginning of each episode, the agent starts with the actual Belgian energy system of 2020 (see Appendix \ref{app:bel_2020}). Moreover, a sample of values, drawn for the uncertain parameters, affects the model for the 2020-2030 time window. This sample will remain valid for the following time windows. In other words, there is only one sample draw per episode (see Figure \ref{fig:ranges_transition}).  Then, the agent takes a set of actions, affecting the environment that feeds back the agent with a new state and a reward. This goes on until the end of the episode. For the new episode, similarly to the \gls{UQ} analysis (see Section \ref{sec:meth:UQ}), the new sample of uncertain parameters for the new episode is drawn following the quasi-random Sobol' sampling technique \cite{bratley2003implementing}.

>>>>

Before testing the optimal policy $\pi^*\left(\bm{a}_n | \bm{o}_n\right)$, the first step consists in assessing the training of the \gls{NN}. For this, numerous episodes are played through the myopic optimization of the transition pathway of Belgium. At the beginning of each episode, the agent starts with the actual Belgian energy system of 2020 (see Appendix \ref{app:bel_2020}). Then, a sample of values, drawn for the uncertain parameters, affects the model for the 2020-2030 time window. This sample will remain valid for the following time windows. In other words, there is only one sample draw per episode (see Figure \ref{fig:ranges_transition}).  Then, the agent takes a set of actions, affecting the environment that feeds back the agent with a new state and a reward. This goes on until the end of the episode. For the new episode, similarly to the \gls{UQ} analysis (see Section \ref{sec:meth:UQ}), the new sample of uncertain parameters for the new episode is drawn following the quasi-random Sobol' sampling technique \cite{bratley2003implementing}.

To reduce the computational burden and maximise the exploration of the transition pathways, the learning phase, also called ``training'', of the agent has been done on the monthly model. Even though averaging time series of end-use demands and renewable productions brings some discrepancies (\ie faster emergence of local \gls{VRES} and smaller electrification of the system), the main advantage of the monthly approach is its computational time, \ie couple of seconds versus about 15 min for the hourly model (see Appendix \ref{app:mo_vs_td}).

### 2024-04-04 chap_atom_mol/chap_atom_mol.tex
Intro post-review DC

``Country specific energy studies are needed as a prerequisite to the decision of following the nuclear route'' \cite{IAEA2008}. Even though, in a country like Belgium, reaching the goal of energy transition will not be a ``winner-takes-all'' situation but rather a combination of solutions, implemented simultaneously \cite{Limpens2020}, this chapter focuses on this atom-molecules dilemma in two ways. First, Section \ref{sec:atom_mol:results_deter} targets the impact of integrating \gls{SMR} from 2040 onward on the whole-energy system, in a deterministic way (\ie considering only nominal values of the parameters). Second, accounting for uncertainties as presented in section \ref{subsec:meth:UQ}, Section \ref{sec:atom_mol:results_uq} will identify the key factors driving higher or lower imports of electrofuels as well as the installation of \gls{SMR}.

>>>>

``Country specific energy studies are needed as a prerequisite to the decision of following the nuclear route'' \cite{IAEA2008}. Even though, in a country like Belgium, reaching the goal of energy transition will not be a ``winner-takes-all'' situation but rather a combination of solutions, implemented simultaneously \cite{Limpens2020}, this chapter focuses on this atom-molecules dilemma in two ways. First, Section \ref{sec:atom_mol:results_deter} targets the impact of integrating \gls{SMR} from 2040 onward on the whole-energy system, in a deterministic way (\ie considering only nominal values of the parameters). Second, accounting for uncertainties as presented in Section \ref{sec:meth:UQ}, Section \ref{sec:atom_mol:results_uq} will identify the key factors driving higher or lower imports of electrofuels as well as the installation of \gls{SMR}.
